---
title: "R Datacamp notes"
author: "Faisal Samin"
date: "03/03/2018"
output: 
  html_notebook:
    number_sections: yes
    theme: cosmo
    # "cerulean", "journal", "flatly", "readable", "spacelab", and "cosmo".
    toc: yes
    toc_depth: 2
    toc_float: yes
    highlight: tango
---


Notes taken during Datacamp course.

# Basics

```{r eval = FALSE}
# Basic arithmetic in R
3 + 4
4 * 2
3 * 6
2 ^ 4
10 %% 3
# Variable assignment
x <- 42
```

There are different types of variables: numerics (floats), integers, boolean values and characters (strings). 

```{r eval = FALSE}
# Check data type with class()
class(5) # numeric
class("Hello World") # character
class(TRUE) # logical
``` 


# Vectors

```{r}
# Vector assignment 
numeric_vector <- c(1, 10, 49)
```

Vectors can contain any types of variables. 

```{r}
# Assign index names to a vector
names(numeric_vector) <- c("a", "b", "c")
```

```{r}
# Vector arithmetic 
A <- c(1, 2, 3, 4)
B <- c(4, 5, 6, 7)
total <- A + B
```

```{r}
# Select values in vector - indexing starts from 1
A[1] # first value
A[c(2,3,4)] # select second, third and fourth values 
A[2:4] # select second, third and fourth values 
A[c('Monday', 'Tuesday')] # seelect columns by name
```

# Matrices

You can construct a matrix in R with the matrix() function.
```{r}
matrix(1:9, byrow = TRUE, nrow = 3)
# The argument byrow indicates that the matrix is filled by the rows. If we want the matrix to be filled by the columns, we just place byrow = FALSE.
# The third argument nrow indicates that the matrix should have three rows.
```

Can also consruct a matrix by combining multiple vectors:
```{r}
new_hope <- c(460.998, 314.4)
empire_strikes <- c(290.475, 247.900)
return_jedi <- c(309.306, 165.8)
star_wars_matrix <- matrix(c(new_hope, empire_strikes, return_jedi), nrow = 3, byrow = TRUE)
```

Naming columns and rows
```{r}
colnames(star_wars_matrix) <- c("US", "non-US")
rownames(star_wars_matrix) <- c("A New Hope", "The Empire Strikes Back", "Return of the Jedi")
```

```{r}
rowSums(star_wars_matrix) # sums of rows 
colSums(star_wars_matrix) # sums of columns
cbind(a,b) # merges matrices and/or vectors together by column
rbind(a,b) # merges matrices and/or vectors together by row
```

Selection of matrix elements

```{r}
my_matrix[1,2] #selects the element at the first row and second column.
my_matrix[1:3,2:4] # results in a matrix with the data on the rows 1, 2, 3 and columns 2, 3, 4.
my_matrix[,1] # selects all elements of the first column.
my_matrix[1,] # selects all elements of the first row.
```

You can also perform arithmetic operations on matrices (+, -, /, *)

# Factors

The term **factor** refers to a statistical data type used to store categorical variables. The difference between a categorical variable and a continuous variable is that a categorical variable can belong to a limited number of categories. A continuous variable, on the other hand, can correspond to an infinite number of values.

It is important that R knows whether it is dealing with a continuous or a categorical variable, as the statistical models you will develop in the future treat both types differently.









